---
title: "Introduction to `SpatialData`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('SpatialData')`"
author: 
  - name: Helena L. Crowell
  - name: Louise Deconinck
  - name: Artür Manukyan
  - name: Dario Righelli
  - name: Estella Dong
  - name: Samuel Gunz
  - name: Yixing E. Dong
  - name: Michael Stadler
  - name: Charlotte Soneson
  - name: Vince Carey
  - name: Hugo Gruson
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to SpatialData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "refs.bib"
---

<style type="text/css">
.smaller {
  font-size: 10px
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
```

# Introduction

The `r BiocStyle::Biocpkg("SpatialData")` package provides an R interface to the 
[SpatialData](https://spatialdata.scverse.org) framework, a unified ecosystem 
for handling spatial omics data. Developed as part of the 
[scverse](https://scverse.org) project [@Virshup2023-scverse], `SpatialData` aims to solve the 
challenges of integrating diverse spatial datasets—including imaging, spatial 
transcriptomics, and proteomics—by employing the 
[OME-NGFF (Next Generation File Format)](https://ngff.openmicroscopy.org) 
standard [@Marconato2025-SpatialData].

The Python implementation and core specifications can be found at the 
[official SpatialData website](https://spatialdata.scverse.org).

## Data Representation in R

The core of the package is the `SpatialData` class, which serves as a container 
for multiple layers of spatial information:

- **Images**: Multiscale, multi-channel raster data, represented as `ZarrArray` objects via the `r BiocStyle::Biocpkg("Rarr")` and `r BiocStyle::Biocpkg("ZarrArray")` packages.
- **Labels**: Segmentation masks (e.g., cell or nuclei boundaries), also stored as `ZarrArray`s.
- **Points**: High-resolution coordinates (e.g., individual transcript molecules), handled as `r BiocStyle::CRANpkg("arrow")` tables or `r BiocStyle::CRANpkg("duckdb")`-backed data frames for efficient out-of-memory processing.
- **Shapes**: Polygonal regions of interest or circular markers, represented similarly to points but with geometric metadata.
- **Tables**: Functional annotations (e.g., gene expression counts), stored as `r BiocStyle::Biocpkg("SingleCellExperiment")` objects.

# Getting Started

First, we load the necessary libraries:

```{r load-libs}
library(SpatialData)
library(SingleCellExperiment)
```

## Reading Data

`SpatialData` objects are typically read from `.zarr` stores. The package provides 
the `readSpatialData()` function to ingest an entire store. For this demonstration, 
we use a toy dataset included in the package.

```{r read-data}
# path to 'spatialdata' .zarr store
zs <- file.path("extdata", "blobs.zarr")
zs <- system.file(zs, package="SpatialData")

# read as 'SpatialData' R object
(sd <- readSpatialData(zs))
```

The output above summarizes the `SpatialData` object, showing the dimension 
of elements in each layer (images, labels, points, shapes, tables), as well 
as the defined coordinate systems and which elements map to each of these.

## Exploring the Object

The `SpatialData` object behaves like a named list of layers. We can inspect 
the names of elements in a given layer:

```{r explore}
imageNames(sd)
shapeNames(sd)
```

We can retrieve a specific element using its name or index:

```{r explore-element}
# retrieve 1st label element
(l <- label(sd))

# access data underlying a point element
p <- point(sd)
as.data.frame(head(p))
```

Elements are S4 objects composed of two key slots:

- `data`: a `list` of `ZarrArray`s for images and labels (representing different 
  multiscale resolutions), or a `duckspatial` data frame for shapes and points.
- `meta`: a `Zattrs` object containing the OME-NGFF metadata retrieved from 
  the `.zattrs` file.

## AnnData Integration with `anndataR`

For spatial omics datasets, functional annotations are typically stored as 
[AnnData](https://anndata.readthedocs.io) objects in Python. 
`SpatialData` uses the `r BiocStyle::Biocpkg("anndataR")` R/Bioconductor 
package [@Deconinck2025-anndataR] to seamlessly read these `.zarr`-backed 
`AnnData` objects into R as `r BiocStyle::Biocpkg("SingleCellExperiment")`.

# Coordinate Systems and Transformations

A key feature of the `SpatialData` framework is its robust handling of 
coordinate systems. Each element can exist in multiple coordinate spaces 
simultaneously, defined by transformations in its metadata.

## Coordinate Transformation Graph

The relationships between different elements and their respective coordinate 
spaces can be complex. `SpatialData` provides the `CTgraph()` and `CTplot()` 
functions to construct and visualize a directed graph of these relationships.

```{r ct-graph, fig.cap="Coordinate transformation graph for the 'blobs' dataset."}
if (requireNamespace("Rgraphviz", quietly=TRUE)) {
    g <- CTgraph(sd)
    CTplot(g)
}
```

In this graph:

- **source nodes** (prefixed with `_`) represent the individual data layers (images, labels, etc.).
- **target nodes** represent the coordinate systems (e.g., `global`, `sequence`).
- **edges** represent the transformations (scale, translation, etc.) required to map an element into a specific space.

## Applying Transformations

The `transform()` function resolves the necessary steps to project an element 
into a target coordinate system by traversing this graph, and applying the 
respective transformation(s):

```{r transform}
# retrieve element & project into 'global'
b <- transform(a <- label(sd), "scale")
do.call(rbind, c(a=extent(a), b=extent(b)))
```

# Spatial Queries

`SpatialData` supports spatial operations, such as cropping by a rectangular
bounding box or polygon, which are propagated across all present layers:

```{r crop}
# bounding box
bb <- list(
    xmin=10, xmax=50, 
    ymin=20, ymax=30)

# subset object
sp <- crop(sd, bb)

# before vs. after
rbind(
    raw=dim(label(sd)),
    box=dim(label(sp))) |> 
    `colnames<-`(c("y", "x"))
```

# Integration with Tables

Tables (SCE objects) link to specific labels or shapes using `region` and 
`instance_key` identifiers, allowing for seamless integration of spatial and 
feature-level information.

```{r tables}
# access annotation
(se <- table(sd))

# annotated region(s)
region(se)
    
# annotated instance(s)
instances(se)
```

# Conclusion

The `SpatialData` package brings the power of the scverse spatial ecosystem to 
R, providing a scalable and interoperable foundation for spatial omics analysis. 
By adhering to the OME-NGFF standard and integrating with established 
Bioconductor objects, it enables seamless cross-platform workflows.

# Session Info {- .smaller}

```{r sessionInfo}
sessionInfo()
```

# References
