## CSAMA

### preamble

```{r load-libs}
library(sf)
library(dplyr)
library(ggplot2)
library(anndataR)
library(scrapper)
library(patchwork)
library(reticulate)
library(SpatialData)
library(SpatialData.data)
library(SpatialData.plot)
library(SingleCellExperiment)
```

### setup

An existing environment that includes Python dependencies can be used via:

```{r eval=FALSE}
# path to Conda binary
bin <- ".../bin/conda" 
options(reticulate.conda_binary=bin)
# activate environment
use_condaenv("<name>")
```

Alternatively, a virtual environment can be created directly from R via:

```{r eval=FALSE}
req <- c("access==1.1.9", ...)
install_python(version=ver <- "3.8.18")
virtualenv_create(envname=env <- "SegTraQ", python=ver, packages=req)
use_virtualenv(env, required=TRUE)
```

```{r load-zarr}
url <- "https://zenodo.org/records/20083116/files/BC_xenium_sdata_lowres.zarr.zip?download=1"
dir <- dirname(zip <- tempfile("sd", fileext=".zarr.zip"))
download.file(url, zip, quiet=TRUE)
fnm <- unzip(zip, exdir=dir)
(sd <- readSpatialData(dirname(fnm[1])))
```

### analysis

We'll start out in Python and load the data into a `spatialdata` object:

```{python load-data}

```

### quality control

### exploratory

#### histopathology

Let's start out by visualization the histopathology image, 
i.e., the hematoxylin and eosin (H\&E) staining:

```{r plt-hne}
cl <- rep(list(c(0, 1/3)), 3) # contrast
plotSpatialData() + plotImage(sd, cl=cl) 
```

We can also zoom into a particular region by specifying a bounding box and
using `crop()` to subset the data accordingly; to do this more efficiently,
we subset the object to contain only images before cropping:

```{r plt-hne-box}
bb <- list(xmin=2000, xmax=4000, ymin=2000, ymax=3000)
sp <- crop(sd["images", ], bb)
plotSpatialData() + plotImage(sp, cl=cl) 
```

#### immunofluorescence

In addition to the H\&E, the dataset also includes four immunofluorescence (IF)
images capturing DAPI (nuclei), E-Cadhering (epithelia), 18S (cytoplasm), and
Vimentin (stromal); the underlying image is thus a 3D array where the first
dimension are channels, the second and third dimensions are height and width.

```{r chs}
img <- image(sd, 2)
cat(channels(img))
cat(dim(img))
```

For consistent visualization, we'll first specify colors to use for each channel:

```{r}
pal <- c("blue", "green", "cyan", "magenta")
chs <- names(pal) <- channels(img)
```

Let's first visualize each channel separately (using the previously defined region):

```{r}
ps <- lapply(chs, \(.) plotSpatialData() + plotImage(sp, 2, ch=., c=pal[.]) + ggtitle(.))
wrap_plots(ps, nrow=2) & theme_void() & theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

Alternatively, we can visualize a four-plex composite image (using the full tissue):

```{r}
plotSpatialData() + plotImage(sd, 2, ch=chs, c=pal[chs])
```

### processing

The `SpatialData` object contains a single `table` element, represented as a
`r BiocStyle::Biocpkg("SingleCellExperiment")`, annotating `cell_boundaries`:

```{r get-tbl}
(se <- tables(sd)[[1]])
```

We can perform standard processing tasks using tools already 
available in R/Bioc, e.g., `r BiocStyle::Biocpkg("scrapper")`:

```{r pro}
# standard 'scrapper' processing:
# - log-library size normalization
# - highly variable gene selection
# - principal component analysis
# - shared-nearest neighbor graph construction
# - community detection using Leiden algorithm
assayNames(se) <- "counts"
se <- normalizeRnaCounts.se(se)
se <- se[, se$sizeFactor > 0]
se <- chooseRnaHvgs.se(se)
se <- runPca.se(se, features=rowData(se)$hvg)
se <- clusterGraph.se(se, method="leiden", resolution=0.5)
table(sd) <- se
```

### downstream

### appendix

::: {.callout-note icon=false, collapse=true}

### session

```{r sess-info}
sessionInfo()
```

:::
