## CSAMA

In this demo, we'll be analyzing 5K-plex Xenium (10x Genomics) 
data on a tissue section from a human breast cancer biopsy, 
derived from [here](https://www.10xgenomics.com/datasets/xenium-prime-ffpe-human-breast-cancer).
These include the following elements:

- two `image`s (histology and immunofluorescence)
- two `shape`s (nuclei and membrane boundaries)
- one `table` (gene $\times$ cell data)
- one `point` (RNA transcripts)

A size-reduced example dataset has been deposited on Zenodo. 
We'll first retrieve the compressed *.zarr* store programatically, 
decompress it, and stash the data's on-disk file path for later:

```{r load-zarr}
# retrieve .zarr store from the web
url <- "https://zenodo.org/records/20083116/files/BC_xenium_sdata_lowres.zarr.zip?download=1"
zip <- tempfile("sd", fileext=".zarr.zip")
download.file(url, zip, quiet=TRUE)
# decompress, stash & view contents
nm <- unzip(zip, exdir=dirname(zip))
zs <- grepv("zarr$", dirname(nm))
cat(basename(zs))
```

```{r show-zarr, echo=FALSE}
htmltools::pre(
    style="max-height:333px; overflow-y:auto;",
    paste(capture.output(fs::dir_tree(zs, recurse=2))[-1], collapse="\n"))
```

### dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(anndataR)
library(scrapper)
library(patchwork)
library(reticulate)
library(SpatialData)
library(SpatialData.data)
library(SpatialData.plot)
library(SingleCellExperiment)
```

### setup

We can instruct our session to use an existing environment with our Python 
dependencies (here, `SegTraQ`) as follows. Note that `reticulate` provides 
similar commands to, e.g., use a virtual (not conda) environemnt instead:

::: {.column-margin}
The environment used here has been set up previously outside of R, 
from the command line, using these commands:
```
conda create -n segtraq python=3.11
conda activate segtraq
pip install segtraq
```
:::

```{r py-env}
# path to conda binary
bin <- "~/software/mambaforge/bin/conda" 
options(reticulate.conda_binary=bin)
# activate environment
use_condaenv("segtraq")
```

### segmentation

```{python, eval=FALSE}
import warnings
warnings.filterwarnings(action="ignore")

import anndata as ad
import matplotlib.pyplot as plt
import segtraq
import spatialdata as sd
import spatialdata_plot

sdata = sd.read_zarr(r.zs)

sdata.shapes["cell_boundaries"].index.name = "cell_id"
sdata.shapes["nucleus_boundaries"].index.name = "cell_id"

st = segtraq.SegTraQ(sdata,
    images_key="morphology_focus", 
    tables_centroid_x_key=None, 
    tables_centroid_y_key=None)
    
st.run_baseline()

adata = st.sdata.tables["table"]
adata
```

```{python, eval=FALSE}
axes = plt.subplots(1, 3, figsize=(14, 5), constrained_layout=True)[1].flatten()

# plot DAPI image
a = st.sdata.pl.render_images("morphology_focus", channel="DAPI")
a.pl.show(ax=axes[0], title="DAPI", coordinate_systems="global")

# plot histology
b = st.sdata.pl.render_images("he_image")
b.pl.show(ax=axes[1], title="H&E", coordinate_systems="global")

# plot overlay of nuclei & cell boundaries colored by cell area
c = st.sdata.pl.render_shapes(
    element="nucleus_boundaries",
    outline_color="black")
c = c.pl.render_shapes(
    element="cell_boundaries",
    outline_color="black")
c.pl.show(ax=axes[2], title="overlay", colorbar=True)

plt.show()
```

### exploratory

Let's load up the data in R, skipping the `table` layer. We then retrieve the 
`table`, including `SegTraQ` results, from Python, and set it to annotate cell
segmentation boundaries using `setTable()`; `ik` specifies the `spatialdata`'s
`instance_key` used to link observations between `table` and spatial element.

```{r}
sd <- readSpatialData(zs)#, tables=FALSE)
# sd <- setTable(sd, 
#     i="cell_boundaries", ik="cell_id",
#     y=py$adata$as_SingleCellExperiment())
sd
```

#### histopathology

Let's start out by visualization the hematoxylin and eosin (H&E) staining.
H\&E provides a high-contrast visualization of tissue microarchitecture and 
cellular morphology through differential staining of nuclear and cytoplasmic 
components. To date, whole-slide H\&E images (WSIs) remain the clinical gold 
standard for cancer diagnosis, including tumor grading, staging, and prognosis.

```{r plt-he}
plotSpatialData() + plotImage(sd)
```

We can also zoom into a particular region by specifying a bounding box and
using `crop()` to subset the data accordingly; to do this more efficiently,
we subset the object to contain only images before cropping:

```{r plt-he-box}
bb <- list(
    xmin=2000, xmax=4000, 
    ymin=2000, ymax=3000)
sp <- crop(sd["images", ], bb)
plotSpatialData() + plotImage(sp) 
```

#### immunofluorescence

In addition to the H\&E, the dataset includes 4 immunofluorescence (IF) markers:

* DAPI (nuclei) marks all cell nuclei and is used for cell segmentation
* ATP1A1/CD45/E-Cadherin (membrane) marks cell boundaries and major lineages
* 18S (cytoplasm) marks the cell body/cytoplasmic RNA and overall cell extent
* αSMA/Vimentin (stroma) marks mesenchymal/stromal tissue, including fibroblasts

```{r if-chs}
img <- image(sd, 2)
as.vector(channels(img))
```

The underlying image is thus a 3D array where the first dimension are channels, 
and the second and third dimensions are height and width; we can verify this
by viewing the `axes()` specified in the *.zattrs* associated with the image:

```{r if-dim}
do.call(rbind, axes(img))
cat(dim(img))
```

Let's visualize the four-plex composite image overlaying these markers;
for consistent visualization, we also specify colors to use for each channel:

```{r plt-if-all}
pal <- c("blue", "green", "cyan", "magenta")
chs <- names(pal) <- channels(img)
plotSpatialData() + plotImage(sd, 2, ch=chs, c=pal[chs])
```

Alternatively, we can visualize each channel separately; 
here, using the previously smaller region:

```{r plt-if-one}
ps <- lapply(chs, \(.) plotSpatialData() + plotImage(sp, 2, ch=., c=pal[.]) + ggtitle(.))
wrap_plots(ps, nrow=2) & theme_void() & theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

We can also zoom in further to inspect how well cell segmentation boundaries 
align with the immunofluorescence signals used to infer them:

```{r eval=FALSE}
# sq <- crop(sd, list(xmin=2500, xmax=3000, ymin=2000, ymax=2500))
# plotSpatialData() +
#     plotImage(sq, 2, ch=chs, c=pal[chs]) +
#     plotShape(sq, fill=NA, col="white", linewidth=1)
```

### processing

The `SpatialData` object contains a single `table` element -- represented as a
`r BiocStyle::Biocpkg("SingleCellExperiment")` -- annotating `cell_boundaries`.
This `table` contains the single-cell level data: a gene $\times$ cell count 
matrix (`assay` slot `"X"`), gene/cell metadata (`row/colData`), as well as
spatial coordinates (`reducedDim` slot `"spatial"`).

```{r get-tbl}
(se <- tables(sd)[[1]])
```

Note that these data originally stem from a Python `anndata` object written 
to `.zarr`, which were read into R using `r BiocStyle::Biocpkg("anndataR")`.
We can therefore perform standard processing tasks using tools available 
in R/Bioconductor, e.g., `r BiocStyle::Biocpkg("scrapper")`, including:

- cell-level quality control
- log-library size normalization
- highly variable gene (HVG) selection
- principal component analysis (PCA)
- shared-nearest neighbor (SNN) graph construction
- community detection using the Leiden algorithm

```{r fil}
se <- tables(sd)[[1]]
assayNames(se) <- "counts"
se <- quickRnaQc.se(se, list())
cpa <- log(se$sum / se$cell_area)
ex <- cpa < median(cpa)-2*mad(cpa)
se$ex <- ex | !se$keep
```

```{r}
#| code-fold: true
sp <- sd
table(sp) <- se
plotSpatialData() + 
    plotShape(sp, col=NA, fill="ex") + 
    scale_fill_manual("exclude", values=c("grey", "red"))
```

```{r pro}
se <- se[, !ex]
se <- normalizeRnaCounts.se(se)
se <- chooseRnaHvgs.se(se)
se <- runPca.se(se, features=rowData(se)$hvg)
se <- clusterGraph.se(se, method="leiden", resolution=0.5)
nk <- nlevels(se$clusters)
```

For convenience, we'll add a few principal components (PCs) as cell metadata:

```{r}
npc <- seq_len(3)
pcs <- reducedDim(se, "PCA")[, npc]
colnames(pcs) <- paste0("PC", npc)
colData(se) <- cbind(colData(se), pcs)
```

Lastly, we'll overwrite the object's original `table` with the processed one:

```{r}
table(sd) <- se

# TODO: temporary fix
shape(sd) <- shape(sd) |> mutate(cell_id=`__index_level_0__`)
shape(sd)@meta$spatialdata_attrs$instance_key <- "cell_id"
shape(sd) <- shape(sd)[match(shape(sd)[[2]], instances(se), nomatch=0)] 
```

```{r plt-xy-clu, message=FALSE}
plotSpatialData() + 
    plotShape(sd, col=NA, fill="clusters") +
    scale_fill_manual(values=hcl.colors(nk, "Spectral")) 
```

### downstream

#### structures

```{r message=FALSE, warning=FALSE}
library(sosta)
library(spatstat.geom)
```

```{r sosta-ppp, warning=FALSE}
xy <- centroids(transform(shape(sd)))
ow <- owin(xrange=range(xy[, 1]), yrange=range(xy[, 2]))
pp <- ppp(x=xy[, 1], y=xy[, 2], window=ow, marks=se$clusters)
```

For now the sosta output is not yet stored in the object. We will plot it 
using `geom_sf()`. For illustration we'll select for the largest polygons.

```{r sosta-roi, message=FALSE}
roi <- reconstructShapeDensity(pp, markSelect=6, dim=400)
roj <- roi[head(order(st_area(roi), decreasing=TRUE)), ]

plotSpatialData() + 
    plotShape(sd, col=NA, fill="clusters") + 
    geom_sf(data=roj, fill=NA, linewidth=1) +
    scale_fill_manual(values=hcl.colors(nk, "Spectral")) 
```

### appendix

::: {.callout-note icon=false, collapse=true}

### session

```{r sess-info}
sessionInfo()
```

:::
