## CSAMA

### preamble

```{r load-libs}
library(sf)
library(dplyr)
library(ggplot2)
library(anndataR)
library(scrapper)
library(patchwork)
library(reticulate)
library(SpatialData)
library(SpatialData.data)
library(SpatialData.plot)
library(SingleCellExperiment)
```

### setup

An existing environment that includes Python dependencies can be used via:

```{r eval=FALSE}
# path to Conda binary
bin <- ".../bin/conda" 
options(reticulate.conda_binary=bin)
# activate environment
use_condaenv("<name>")
```

Alternatively, a virtual environment can be created directly from R via:

```{r eval=FALSE}
req <- c("access==1.1.9", ...)
install_python(version=ver <- "3.8.18")
virtualenv_create(envname=env <- "SegTraQ", python=ver, packages=req)
use_virtualenv(env, required=TRUE)
```

```{r load-zarr}
url <- "https://zenodo.org/records/20083116/files/BC_xenium_sdata_lowres.zarr.zip?download=1"
dir <- dirname(zip <- tempfile("sd", fileext=".zarr.zip"))
download.file(url, zip, quiet=TRUE)
fnm <- unzip(zip, exdir=dir)
(sd <- readSpatialData(dirname(fnm[1])))
```

```{r}

```

### analysis

We'll start out in Python and load the data into a `spatialdata` object:

```{python load-data}

```

### quality control

### processing

The `SpatialData` object contains a single `table` element, represented as a
`r BiocStyle::Biocpkg("SingleCellExperiment")`, annotating `cell_boundaries`:

```{r get-tbl}
(se <- tables(sd)[[1]])
```

We can perform standard processing tasks using tools already 
available in R/Bioc, e.g., `r BiocStyle::Biocpkg("scrapper")`:

```{r pro}
# standard 'scrapper' processing:
# - log-library size normalization
# - highly variable gene selection
# - principal component analysis
# - shared-nearest neighbor graph construction
# - community detection using Leiden algorithm
assayNames(se) <- "counts"
se <- normalizeRnaCounts.se(se)
se <- chooseRnaHvgs.se(se, top=2e3)
se <- runPca.se(se, features=rowData(se)$hvg)
se <- clusterGraph.se(se, method="leiden", resolution=0.5)
```

### downstream

Let's start out by visualization the histopathology image, 
i.e., the hematoxylin and eosin (H\&E) staining:

```{r plt-hne}
cl <- rep(list(c(0, 1/3)), 3) # contrast
plotSpatialData() + plotImage(sd, cl=cl) 
```

### appendix

::: {.callout-note icon=false, collapse=true}

### session

```{r sess-info}
sessionInfo()
```

:::
